---
title: "Principle of functional enrichment"
author: "Jacques van Helden & Yvon Mbouamboua"
date: "`r Sys.Date()`"
always_allow_html: yes
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    highlight: zenburn
    self_contained: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  word_document:
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: no
    smaller: yes
    theme: cerulean
    toc: yes
    toc_float: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise


```{r parameters}
m <- 3000 ## Number of annotated genes for the funcitonal class
N <- 30000 # Total number of genes with some annotation in the GO catalog
k <- 50 # size of the selection = number of genes associated to the disease
x <- 10 # Number of genes of interest annotated for the current functional class

X <- 0:k #All possible values for the size of the intersection

```


The principle of enrichment analysis is to detect significant over-representation of funcional classes among a set of genes of interest (**GOI**). 

We want to analyse the functional enrichment of a set of $k = `r k`$ genes associated to a disease of interest. We analyse the funcitonal annotations of these genes in a reference catalog (for example the Gene Ontology catalog, or a Human disease catalog). 

We analyse  one by one all the classes of the catalog.  

Let us imagine that a given functional class (e.g. lung diseases) contains $m=`r m`$ genes ("marked" genes), and that they include $x = `r x`$ of our $k = 50$ genes of interest.



## The hypergeometric distribution


```{r hypergeometric_distribution}

## Hypergeometric density (probability to observe exactly x successes)
hyper.dens <- dhyper(x = X -1, m = m, n = N - m, k = k)

## Hypergeometric p-value: probability to observe at least x successes
hyper.pval <- phyper(q = X -1, m = m, n = N - m, k = k, lower.tail = FALSE)

## Plot the distribution 
plot(X, hyper.dens, main = "Hypergeometric density", type = "h", lwd = 3, col ="blue", las = 1,
     xlab = "X = number of successes", ylab = "P(X = x)")

arrows(x0 = x, 
       y0 = (hyper.dens[X == x]) + 0.05, 
       x1 = x, 
       y1 = (hyper.dens[X == x]) * 1.5, 
       length = 0.1, angle = 30, lwd = 2, col = "red")
text(x = x, 
     y = (hyper.dens[X == x]) + 0.05, 
     pos = 4, 
     labels = paste(sep = "\n", 
                    paste("x =", x), 
                    paste("P(X = x) =", signif(digits = 2, hyper.dens[X == x])),
                    paste("pval = P(X >= x) =", signif(digits = 2, hyper.pval[X == x]))), 
     col = "red")

lines(X[X >= x], 
      hyper.dens[X >= x], type = "h", 
      lwd = 3, col ="red")
```

## Binomial approximation


An alternative to the hypergeomztric is to estiamte the prior probability ($p$) to draw by chance a gene that belongs to the functional class, and use it as probability of success the binomial distribution. 

```{r binomial}

## Prior probability to pick up a marked gene in the universe
p.success <- m / N
  
## Binomial density (probability to observe exactly x successes)
binom.dens <- dbinom(x = X -1, size = k, prob = p.success)

## Binomial p-value: probability to observe at least x successes
binom.pval <- pbinom(q = X -1, size = k, prob = p.success, lower.tail = FALSE)

## Plot the distribution 
plot(X, binom.dens, main = "Binomial density", type = "h", lwd = 3, col ="blue", las = 1,
     xlab = "X = number of successes", ylab = "P(X = x)")

arrows(x0 = x, 
       y0 = (binom.dens[X == x]) + 0.05, 
       x1 = x, 
       y1 = (binom.dens[X == x]) * 1.5, 
       length = 0.1, angle = 30, lwd = 2, col = "red")
text(x = x, 
     y = (binom.dens[X == x]) + 0.05, 
     pos = 4, 
     labels = paste(sep = "\n", 
                    paste("x =", x), 
                    paste("P(X = x) =", signif(digits = 2, binom.dens[X == x])),
                    paste("pval = P(X >= x) =", signif(digits = 2, binom.pval[X == x]))), 
     col = "red")

lines(X[X >= x], 
      binom.dens[X >= x], type = "h", 
      lwd = 3, col ="red")
  
```



